Topological Equivariant Artist Model for Visualization Library Architecture

### 2023-02-01 paper/main.tex
pulled up topolology to related work topological structure

There is an assumption in visualization that the connectivity of the elements in a dataset are preserved in the visual representation; roughly topology provides a method for encoding this connectivity \cite{wilkinsonGrammarGraphics2005}. Bertin gives the examples of discrete data values should be represented by discrete entities such as points, quantitative continuous data by lines, and surfaces through area marks \cite{bertinSemiologyGraphicsDiagrams2011}. In \autoref{fig:related-work:continuity:ktypes} the records of the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} weather station table can be considered disconnected from each other and therefore have a 0D continuity. This means each distinct row can be reduced to a discrete point and each row will be encoded as a discrete visual element, such as a standalone scatter point. Each station can also be considered a time series of weather observation since temperature and pressure are sampled from a 1D continuous measurement space. The connectivity of these observations can be modeled as an interval and the data is encoded as a line since that visual element is also 1D continuous. The measurements at set of stations on any given day is a sparse sampling of a 2D geospatial grid and the connectivity can be modeled as a 2D plane. Instead of interpolation, which is how continuity was preserved in the timeseries plot, in this instance the connectivity is preserved by plotting the stations in the map such that their relative distances are preserved. Encoding connectivity using the formalism of topological spaces allows us to encode the connectivity of the data in a uniform manner such that we can then generalize the preservation of that connectivity in the visual representation rather than constructing a condition specifically for each type of topological structure.

@@

\subsubsection{Topological Structure: Base Space \dbase}
\label{sec:atct:fb:base}
We encode the topological structure of the data as the \textcolor{base}{base space} \dbase\ of the fiber bundle since the base space acts as indexing space where every point in the base space has a corresponding fiber space $\pi^{-1}(\dbasepoint) = \dfiber_{\dbasepoint}$. The \textcolor{base}{base space} \dbasec\ is a topological space $(\dbase,\mathcal{T}_{\dbase})$, which means it  consists of the set of points ${\dbasepoint \in \dbase}$ and topology $\mathcal{T}_{\dbase}$ \cite{munkresElementsAlgebraicTopology1984}. A topology $\dbase \coloneqq (\dbase,\mathcal{T}_{\dbase})$ is an axiomatic way of defining a topological structure\cite{bradleyTopologyCategoricalApproach2020}  as a collection of subsets, called open sets \footnote{Open sets (open subsets) are a generalization of open intervals to n dimensional spaces. For example, an open ball is the set of points inside the ball and excludes points on the surface of the ball. \cite{weissteinOpenSet,bradleyTopologyVsTopology}}, of that structure. 
\begin{definition}
A \textbf{topology} $\mathcal{T}$ on a space $X$ is a collection of open sets \openset\ of $X$. This collection has the properties that the empty set $\varnothing$ and $\dbase$ are in $\mathcal{T}$, the union of open sets in $\mathcal{T}$ is also in $\mathcal{T}$, and any finite intersection of open sets in $\mathcal{T}$ is also in $\mathscr{T}$. \cite{bradleyTopologyCategoricalApproach2020}
\end{definition}
We propose that the topology on \dbase\ can act as a mathematical abstraction for the \textcolor{base}{indexing space} of a dataset because the properties of a topology are the same as would be expected of an index space, namely that the space has a mathematical structure, that the space can be subsetted, and that continuity is preserved when sets of indices are merged. For example, in \autoref{fig:related-work:continuity:ktypes}, a topology on the line would break the line up into subsets, and those subsets could only be joined in ways where they cover the line in the same order in which it was broken up. The base space of a fiber bundle is a quotient topology\cite{QuotientSpaceTopology2020,munkresElementsAlgebraicTopology1984}, meaning that it divides the topological space into the smallest possible (largest number of) open sets such that $\pi$ remains a continuous function. This means that the topology can be defined to have a resolution equal to the number of indicies in a dataset or at whatever resolution of continuois function is required for a given task.

>>>>

The connectivity and relative positioning of elements in a dataset is broadly its topology\cite{wilkinsonGrammarGraphics2005} and the expectation that it be be reflected in the visual representation is present in even the earliest maps \cite{friendlyBriefHistoryData2008} and is exemplified in Bertin's recommendation that discrete data values should be represented by discrete entities such as points, quantitative continuous data by lines, and surfaces through area marks \cite{bertinSemiologyGraphicsDiagrams2011}. Encoding this point, interval, and surface, connectivity as mathematical topologies provides us with a consistent way to describe and discuss these and other types of connectivity and the topological structure of the corresponding visualizations.

\begin{definition}
A \textbf{topology} $\mathcal{T}$ on a space $X$ is a colletion of open sets\footnote{Open sets (open subsets) are a generalization of open intervals to n dimensional spaces. For example, an open ball is the set of points inside the ball and excludes points on the surface of the ball. \cite{weissteinOpenSet,bradleyTopologyVsTopology}} \openset\ of $X$. This collection has the properties that the empty set $\varnothing$ and $\dbase$ are in $\mathcal{T}$, the union of open sets in $\mathcal{T}$ is also in $\mathcal{T}$, and any finite intersection of open sets in $\mathcal{T}$ is also in $\mathscr{T}$. \cite{bradleyTopologyCategoricalApproach2020}
\end{definition}

 In \autoref{fig:related-work:continuity:ktypes} the records of the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} weather station table can be considered disconnected from each other and therefore have a 0D continuity. This means each distinct row can be reduced to a discrete point and each row will be encoded as a discrete visual element, such as a standalone scatter point. Each station can also be considered a time series of weather observation since temperature and pressure are sampled from a 1D continuous measurement space. The connectivity of these observations can be modeled as an interval and the data is encoded as a line since that visual element is also 1D continuous. The measurements at set of stations on any given day is a sparse sampling of a 2D geospatial grid and the connectivity can be modeled as a 2D plane. Instead of interpolation, which is how continuity was preserved in the timeseries plot, in this instance the connectivity is preserved by plotting the stations in the map such that their relative distances are preserved.

\note{pull definition of topology up here and also maybe homotopy type/homomorphic spaces in the next paragraph-it'll make things less handwavy }

Encoding connectivity using the formalism of topological spaces allows us to encode the connectivity of the data in a uniform manner such that we can then generalize the preservation of that connectivity in the visual representation rather than constructing a condition specifically for each type of topological structure.

@@

\subsubsection{Topological Structure: Base Space \dbase}
\label{sec:atct:fb:base}
We encode the topological structure of the data as the \textcolor{base}{base space} \dbase\ of the fiber bundle since the base space acts as indexing space where every point in the base space has a corresponding fiber space $\pi^{-1}(\dbasepoint) = \dfiber_{\dbasepoint}$. The \textcolor{base}{base space} \dbasec\ is a topological space $(\dbase,\mathcal{T}_{\dbase})$, which means it  consists of the set of points ${\dbasepoint \in \dbase}$ and topology $\mathcal{T}_{\dbase}$. We propose that the topology on \dbase\ can act as a mathematical abstraction for the \textcolor{base}{indexing space} of a dataset because the properties of a topology are the same as would be expected of an index space, namely that the space has a mathematical structure, that the space can be subsetted, and that continuity is preserved when sets of indices are merged. For example, in \autoref{fig:related-work:continuity:ktypes}, a topology on the line would break the line up into subsets, and those subsets could only be joined in ways where they cover the line in the same order in which it was broken up. The base space of a fiber bundle is a quotient topology\cite{QuotientSpaceTopology2020,munkresElementsAlgebraicTopology1984}, meaning that it divides the topological space into the smallest possible (largest number of) open sets such that $\pi$ remains a continuous function. This means that the topology can be defined to have a resolution equal to the number of indicies in a dataset or at whatever resolution of continuois function is required for a given task.
